Peptide retention time prediction

### 2015-10-06 response.tex

The authors give a good overview of the existing methods to predict the retention time.  However, the description of the methods is sometimes too brief, and slightly biased towards their own developed methods. Extending the descriptions of the methods that are not developed by the authors should be considered. The applications and discussion section are an added value to the paper.

\ereview

We would like to thank the reviewer for the positive feedback. We have revised the sections were the other methods contribution is discussed

\breview
Additional comments:
\begin{itemize}
\item Legend of Figure 2 overlaps with text on page 3
\item Table 1 is not fully visible; additionally in the table ``Elude'' is written, while in the text ``ELUDE'' is used
\end{itemize}
\ereview
We have corrected this.
\breview
\begin{itemize}

@@

\ereview

Thank you for highlighting these typos. We have now corrected them.

\section*{Reviewer \#2}

The reviewer feels the paper’s target audience isn’t clear. As such,
the overall sensation is that the paper touches several subjects
lightly but fails in going in detail into any of them. As it is right
now, this paper is not practical enough for a mass spectrometrist, and
not technical enough for a machine learning scientist. The reviewer
feels it is necessary to define an audience and focus the paper to
this audience’s expertise.

Specific Comments:

As it was mentioned before, the main concern of the reviewer is the lack of focus in the paper. As such the main point of the review is to clarify who is the audience the paper is meant for and delve deeper into what is relevant for them.

We feel that while the reviewer is right in that this manuscript
targets more than one audience, and that we due to brevity reasons
have to skip some details. However, we also feel that this is the point of a review, to give an overview for whoever wants to get an introduction to a subject. The details can be found in the publications we refer to. Regardless, we updated the text to better fit into a category of reader that we would define as a mass spectrometrist with a fair understanding of computational methods that is interested in knowing were the field of retention time predictions stands today.

Other than that, there are some points worth mentioning. They will be here presented in the order they appear on the manuscript.

On section 2, it is written ``so-called linear gradients''. Why say ``so-called''? This expression gives the idea that the gradients in reality are not linear, but this isn’t further developed.

\ereview
We replaced the word so-called with setting the words linear gradient in italics, to indicate that a explanation of the term will follow. 
\breview

The reproducibility of RPLC is a big issue. Not only does a column get degraded with use, but there is also documented variability in products from different manufacturers. In this section it is also noted how the retention time depends on several experimental settings – ``variety of parameters such as column dimensions, the composition of stationary and mobile phase, temperature or gradient slope''. So we have several factors independent of the peptide sequence that influence greatly the retention time of a given peptide. In the following section it is said that ``given a chromatography setup, the 
sequence of a peptide should provide virtually all the information needed to predict its retention time''; however, given that there is so much variability in the chromatography setup in different labs and different experiments, we feel that it should be emphasized how big of an issue this is in retention time prediction and note efforts made in addressing this.

\ereview
We added a qualifying section to this section

``However, in practice such predictive models is not as straight
forward as it might seem. For instance, one important part of such a
model would be the peptides three dimensional conformation. Although
the current structure prediction methods are quite accurate in
predicting three dimensional conformation of short amino acid chains,
such methods will not be equipped with the flexibility to model the
various conditions of a chromatographic experiment. For instance, many
peptides are likely changing their confirmation as a function of the
amount of organic solvent in the mobile phase, an effect that is
likely to have an important contribution to each peptides retention
time. Hence, most efforts to predict retention time have moved into
treating the chromatographic process as a black box, that is a system
that is only viewed in terms of its inputs an outputs.''

There are some issues with the caption for Fig. 2. It is also not clear if there are points that fall on the regression line. 
\ereview

We changed the formating of the manuscript so that the caption of Figure2 should stand clear. There is no regression line in Fig. 2.

Besides, this figure is not representative of a complex sample or of the reproducibility (or variability) in experiments between different labs, or even different machines. As such the figure can be misleading in terms of evaluating the reproducibility of a shotgun proteomics experiment.

We added a qualifying sentence to the caption of the Figure: ``It
should be noted that this relatively high reproducibility normally
only appears between runs under near to identical experiments, and do
not carry over into other experimental conditions or instrumentation.''

Across section 3, there is a lack of evaluation of the prediction methods. In fact at a certain point it is said that one is ``good'', one ``excellent'', and another that has ``lower'' accuracy followed by one that is ``reasonable''. It is inferred that if these qualitative comparisons are made by the authors, then there must be some accuracy measure available. It would be preferable to have a table with these values.
\ereview

We removed all these rather sloppy unqualified references to methods accuracy.

Table 1 falls out of the page.
\ereview
We modified the formating of the table.
\breview

In section 3.1, the last paragraph doesn’t seem relevant to the section it is included in. Maybe it would make more sense to include it in the applications section.
\ereview

We did not feel that the paragraph sat better in the application section and letted the sentence remain as is.

In section 3.2, it feels relevant to introduce the approach ELUDE takes regarding retention indexes. Even if they obtained via, are used as features for, a machine learning approach, it is still relevant to this section.
\ereview

We added a couple of sentences describing this.
\breview

Section 3.3 is rather brief. The reviewer feels there should be more information on these methods (or none at all).
\ereview
\breview

On sections 3.1, 3.2 and 3.3 there is no evaluation of any of the introduced methods. In section 3.4 we see the sort of qualitative comparisons mentioned above. We note that there is no mention of how the performance of ELUDE compares to the other methods, even if qualitatively.
\ereview
\breview

If the proposed audience to this paper are scientists applying machine learning methods to retention time prediction, more information on the features used and the methods themselves would be relevant. 
\ereview
\breview

In the discussion the reviewer feels that it should be discussed not only how retention time prediction can be improved, but also the limitations of how accurate predictions can be, given the issues mentioned above. Another relevant question is how much more relevant do these predictions need be given the applications enumerated in section 5.

### 2015-10-06 retention.tex
have to leave

% say something about gradient and washing 
Since the elution of the peptides is determined by the proportion of
solvent B present at each time point, the scheme used to add solvent B
is an essential aspect of any RPLC experiment. Currently, the most
commonly used approach is based on so-called linear gradients, where
the increase in solvent B follows a linear function. Most of the
peptides elute before the percentage of solvent B reaches 40\%, and
thus gradients often go from 2 to 40\% B. A high increase in
percentage B at the end of the run ensures that all the remaining
compounds are washed off from the column.

@@

Liquid chromatography gives reproducible results. Except for small
shifts from run to run, the same peptide elutes at about the same time
when a sample is reanalyzed under identical experimental conditions
(Figure~\ref{fig:repr}). This suggests that peptides engage in
well-defined interactions during separation, which depend only on the
experimental setup employed and the properties of the peptides
themselves. However, the properties of the peptides are determined to
a large extent by their amino-acid sequence. As a consequence, given a
chromatography setup, the sequence of a peptide should provide
virtually all the information needed to predict its retention time.
However, in practice this is not as straight forward as it might seem, as it for instance is difficult to predict a peptide's three dimensional structure from its amino acid sequence.

%\begin{wrapfigure}{r}{.45\textwidth}
%\vspace{-10pt}
\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
    time.} Two identical shotgun proteomics experiments were carried
  out. The retention times of the peptides confidently identified in
  both runs are displayed.}
\end{figure}
%\vspace{-25pt}
%\end{wrapfigure}

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are illustrated in
Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
of the most popular retention time prediction methods from each
category.

@@

\begin{table}
 \caption{{\bf Retention time prediction methods.} We
   give a few of the most commonly used retention time prediction methods available.}
 \vspace{.2cm}
 \label{tab:rtmethods}
 \begin{tabular}{lll}

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
% ANN & Link? & \cite{petritis2003,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

{\sc Elude} & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
iRT-calculator & http://www.biognosys.ch/products/irt-calculator.html
& \cite{irt} \\
\hline
 \end{tabular} \\
\end{table}

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}.

@@

One of the first machine learning approaches for retention time
prediction was based on an Artificial Neural Network
(ANN) \cite{petritis2003}. Initially, each peptide was described by 20
features giving the number of each of the 20 amino acid residues
present in the sequence. The network was trained using $\sim$7000
peptides with known retention times and evaluated using a set of
$\sim$5000 peptides originating from another organism. The algorithm
reported good accuracy, especially taking into account that it only
used information about the amino acid composition of the peptides.

In a follow-up work, the method was extended by representing each
peptide by 1052 instead of 20 features
\cite{petritis2006improved}. The new descriptors encoded the identity
of each amino acid at each position in a sequence, as well as the
length and the hydrophobic moment of the peptide. The new neural
network was trained on more than 300,000 peptide sequences. To obtain
such a large number of peptides with known retention times, a strategy
to normalize elution times was developed. This allowed pooling of
peptides identified in more than 12,000 shotgun experiments. The
trained algorithm reported excellent performance for the 1303 peptides
on which it was evaluated.

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracy. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

@@

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide a prediction precision comparable to the small shifts
observed when comparing retention time of peptides from two technical
replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to reach the desired level of accuracy.

>>>>

% say something about gradient and washing 
Since the elution of the peptides is determined by the proportion of
solvent B present at each time point, the scheme used to add solvent B
is an essential aspect of any RPLC experiment. Currently, the most
commonly used approach is based on {\em linear gradients}, where
the increase in solvent B follows a linear function. Most of the
peptides elute before the percentage of solvent B reaches 40\%, and
thus gradients often go from 2 to 40\% B. A high increase in
percentage B at the end of the run ensures that all the remaining
compounds are washed off from the column.

@@

Liquid chromatography gives reproducible results. Except for small
shifts from run to run, the same peptide elutes at about the same time
when a sample is reanalyzed under identical experimental conditions
(Figure~\ref{fig:repr}). This suggests that peptides engage in
well-defined interactions during separation, which depend only on the
experimental setup employed and the properties of the peptides
themselves. However, the properties of the peptides are determined to
a large extent by their amino-acid sequence. As a consequence, given a
chromatography setup, the sequence of a peptide should provide
virtually all the information needed to predict its retention time.

\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
    time.} Two identical shotgun proteomics experiments were carried
  out. The retention times of the peptides confidently identified in
  both runs are displayed. It should be noted that this relatively high reproducibility normally only appears between runs under near to identical experiments, and do not carry over into other experimental conditions or instrumentation.}
\end{figure}

However, in practice such predictive models is not as straight forward
as it might seem. For instance, one important part of such a model
would be the peptides three dimensional conformation. Although the
current structure prediction methods are quite accurate in predicting
three dimensional conformation of short amino acid chains, such
methods will not be equipped with the flexibility to model the various
conditions of a chromatographic experiment. For instance, many
peptides are likely changing their confirmation as a function of the
amount of organic solvent in the mobile phase, an effect that is
likely to have an important contribution to each peptides retention
time. Hence, most efforts to predict retention time have moved into
treating the chromatographic process as a black box, that is a system
that is only viewed in terms of its inputs an outputs.

Following these observations,a variety of retention time prediction methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are illustrated in
Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
of the most popular retention time prediction methods from each
category.

@@

\begin{table}
 \caption{{\bf Retention time prediction methods.} We
   give a few of the most commonly used retention time prediction methods available.}
 \vspace{.2cm}
 \label{tab:rtmethods}
 \begin{tabular}{lp{9cm}l}

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
% ANN & Link? & \cite{petritis2003,petritis2006improved} \\
RTPredict &  \url{ http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
RTModel &\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \cite{rtpredict, rtpredictImproved} \\

{\sc Elude} & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
iRT-calculator & \url{http://www.biognosys.ch/products/irt-calculator.html} & \cite{irt} \\
\hline
 \end{tabular} \\
\end{table}

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
Elude} \cite{elude1,elude2} estimates retention time indices for each
dataset as hand. In these cases the indeces are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

One of the first machine learning approaches for retention time
prediction was based on an Artificial Neural Network
(ANN) \cite{petritis2003}. Initially, each peptide was described by 20
features giving the number of each of the 20 amino acid residues
present in the sequence. The network was trained using $\sim$7000
peptides with known retention times and evaluated using a set of
$\sim$5000 peptides originating from another organism.

In a follow-up work, the method was extended by representing each
peptide by 1052 instead of 20 features
\cite{petritis2006improved}. The new descriptors encoded the identity
of each amino acid at each position in a sequence, as well as the
length and the hydrophobic moment of the peptide. The new neural
network was trained on more than 300,000 peptide sequences. To obtain
such a large number of peptides with known retention times, a strategy
to normalize elution times was developed. This allowed pooling of
peptides identified in more than 12,000 shotgun experiments. The
trained algorithm reported better performance for the 1303 peptides
on which it was evaluated, than its predecesor.

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method using a customized
kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided an algorithm that required as little as 50 peptides
for training.

@@

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide a prediction precision comparable to the small shifts
observed when comparing retention time of peptides from two technical
replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to even understand the error sources in the process.
